Almost orthogonal vectors

### 2014-11-27 project3/paper/paper.tex
More changes

\section{Introduction}
Consider a collection of $N$ unit vectors $v_1, \ldots, v_N$ in $\mathbb R^n$. Define $\epsilon$ as
\[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
For various $n$ and $N$, how do we minimize $\epsilon$?

@@

In this paper, we shall see that, for $n=2$, the minimum value of $\epsilon$ will be given by 
\[\epsilon=\cos^2\pi/N.\] 
We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ will be 
\[\epsilon=\frac{1}{n^2},\] 
with the $n+1$ vectors pointing to the vertices of the $n$-simplex.

@@

Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be the reverses of $v_1, v_2, v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because if the vector $\mathcal{R}(v)$ is the reverse of the vector $v$, then $|v \cdot v'| = |\mathcal{R}(v) \cdot v'|$ for some arbitrary vector $v'$.

Without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is the reverse of $x_1$, we see that $x_4 = [-1, 0]^T$. Also, without loss of generality assume that $x_2$ and $x_3$ are above the $x$-axis, and that $x_2$ is to the right of $x_3$.

@@

@@

Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$ -- if any vector $x_i$ were not above the $x$-axis, then we could just consider its reverse.

Let us define $\theta_i$ as the angle between the vectors $x_i$ and $x_{i+1}$. Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$. If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$.

Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$.
\end{proof}

@@

        \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
        \node at (-4.5, 1.76) {\text{$x'_2 = [\cos \pi/5,\sin \pi/5]$}};
        \node at (3.43, -3) {\text{$x_3 = [\cos 2\pi/5,\sin 2\pi/5]$}};
         \node at (-4.5, -1.76) {\text{$x_5 = [\cos 4\pi/5,\sin 4\pi/5]$}};
        \node at (3.43, 3) {\text{$x'_4 = [\cos 3\pi/5,\sin 3\pi/5]$}};

@@

Figure $3$ shows an optimum configuration for $n=2$ and $N=5$. Note by taking reflections of $x_2$ and $x_4$ instead, we get five vectors that form a regular $5$-gon -- this can be seen in Figure $4$. This can be done for all odd $N$; unfortunately, even $N$ do not produce such nice polygons.

@@

\begin{definition}
For a vector $v\in\R^n$ the norm of $v$ is the positive value of $||v||$ given by the equation
$$
||v||^2=v\cdot v.
$$
\end{definition}
\begin{theorem}[Parallelogram Law]
For vectors $v_1,\cdots,v_n\in\R^n$, the following identity holds
$$
||v_1+\cdots+v_n||^2=\displaystyle\sum_{i=1}^n||v_i||^2+2\displaystyle\sum_{1\le i<j\le n}v_i\cdot v_j.
$$
\proof
First consider the following identity from the definition of the norm for vectors $u,v\in\R^n$:
$$
||u+v||^2=(u+v)\cdot (u+v)=u\cdot u+u\cdot v+ v\cdot u + v\cdot v=||u||^2+||v||^2+2u\cdot v.
$$
Now if we let $u=v_1$ and $v= v_2+\cdots+v_n$ we get 
$$
||v_1+v+2+\cdots+v_n||^2=||v_1||^2+||v_2+\cdots +v_n||^2+2\sum_{i=2}^n v_1\cdot v_i.
$$
If we recurse on $||v_2+\cdots +v_n||^2$ using induction we get the desired result. 
\end{theorem}

\section{$(n,n+1)$}
Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$ in this case.

@@

\begin{figure}[!h]
    \centering
    \begin{tikzpicture}[yscale=-1] 
        % x-axis
        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
        % y-axis
        \draw [thick,->] (0, 4.5) -- (0, -4.5);
        % origin label
        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
        % x-axis label
        \node at (4.5, 0.5) {\text{$x$}};
        % y-axis label
        \node at (0, -5) {\text{$y$}};
        % circle
        \draw (0,0) circle (3cm);
        \draw (3,0)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, 2.59)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);

@@

        % \draw [thick,-,blue, dotted] (-3, 0) -- (0, 0);
        % \draw [thick,-,blue, dotted] (1.5, -2.59) -- (0, 0);
        % \draw [thick,-,blue, dotted] (1.5, 2.59) -- (0, 0);
        \draw [thick,-,blue] (3, 0) -- (0, 0);
        \draw [thick,-,blue] (-1.5, 2.59) -- (0, 0);
        \draw [thick,-,blue] (-1.5, -2.59) -- (0, 0);
    \end{tikzpicture}
    \caption{A 2-dimensional simplex drawn in $\R^3$ on the left, and then centered at the origin in $\R^2$ on the right.}
\end{figure}
\missingfigure{Make a sketch of a 2d simplex in $\R^3$}

@@

We must first describe the vectors that make up the origin centered simplex. To center our simplex at the origin we take each vector in the original simplex, and subtract from it the centroid of the original simplex. The centroid of the original simplex takes the form $(\frac{1}{n},\frac{1}{n},\cdots,\frac{1}{N})$, so the points of the new simplex take the form
$$
\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
If we renormalize these vectors we get 
$$
\sqrt{\frac{N}{N-1}}\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
If we take the dot product of any two such vectors we get
$$
\frac{N}{N-1}\left(-2\cdot\frac{N-1}{N^2}+(N-2)\cdot\frac{1}{N^2}\right)=\frac{N}{N-1}\cdot-\frac{1}{N}=-\frac{1}{N-1}=-\frac{1}{n}.
$$
So an upper bound on the value of $\delta$ for all configurations of $n+1$ vectors in $n+1$ dimensional space is $\delta\le\frac{-1}{n}$.

@@

$$
||v_1+\cdots+v_{n+1}||^2=\displaystyle\sum_{i=1}^{n+1}||v_i||^2+2\displaystyle\sum_{1\le i<j\le n+1}v_i\cdot v_j.
$$
In our case we know three things. First, since the vectors are arranged in a simplex, $v_1+\cdots+v_{n+1}=0$. Second, $v_i\cdot v_j\le\delta$ for any $i<j$. Third, $||v_i||=1$. Therefore 
$$
0\le n+1+n(n+1)\delta.
$$
This implies that $\delta\ge -\frac{1}{n}$ as desired. Now we see that $\delta=-\frac{1}{n}$, so $|\delta|=\frac{1}{n}$, and therefore $\epsilon=\frac{1}{n^2}$.

\section{N>n+1}

We conjecture that in cases where $N>n+1$ we can never have an arrangement of the vectors where all values of $|v_i\cdot v_j|^2$ are the same.

>>>>

\section{Introduction}
Consider a collection of $N$ unit vectors $v_1, \ldots, v_N$ in $\mathbb R^n$. Define $\epsilon$ as
\[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
In this paper, we consider the problem of minimizing $\epsilon$ for various $n$ and $N$.

@@

In this paper, we shall see that, for $n=2$, the minimum value of $\epsilon$ will be given by 
\[\epsilon=\cos^2\pi/N.\]

We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ will be 
\[\epsilon=\frac{1}{n^2},\] 
with the $n+1$ vectors pointing to the vertices of the $n$-simplex.

@@

Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be $-v_1, -v_2, -v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because $|v \cdot v'| = |(-v) \cdot v'|$ for some arbitrary vector $v'$.

Now, without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is $-x_1$, we see that $x_4 = [-1, 0]^T$. Also, without loss of generality let us assume that $x_2$ and $x_3$ are above the $x$-axis, and that $x_2$ is to the right of $x_3$.

@@

Given the above result for $N=3$, we attempt to generalize to any integer $N$ in the following theorem.

@@

Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$ -- if any vector $x_i$ were not above the $x$-axis, then we could just consider $-x_i$ instead.

Let us define $\theta_i$ as the angle between the vectors $x_i$ and $x_{i+1}$. Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$. If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$ by the above Lemma.

Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$ (again making use of the fact that the cosine function is decreasing).
\end{proof}

@@

        \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
        \node at (-4.5, 1.76) {\text{$x'_2 = [\cos 6\pi/5,\sin 6\pi/5]$}};
        \node at (3.43, -3) {\text{$x_3 = [\cos 2\pi/5,\sin 2\pi/5]$}};
         \node at (-4.5, -1.76) {\text{$x_5 = [\cos 4\pi/5,\sin 4\pi/5]$}};
        \node at (3.43, 3) {\text{$x'_4 = [\cos 8\pi/5,\sin 8\pi/5]$}};

@@

Figure $3$ shows an optimum configuration for $n=2$ and $N=5$. Note by taking reflections of $x_2$ and $x_4$ in Figure $3$, we get five vectors that form a regular $5$-gon -- this can be seen in Figure $4$. A similar transformation can be done for all odd $N$; unfortunately, even $N$ cannot produce regular polygons.

@@

\begin{definition}
For a vector $v\in\R^n$ the norm of $v$ is the positive value of $||v||$ given by the equation
$$
||v||^2=v\cdot v.
$$
\end{definition}
\begin{theorem}[Parallelogram Law]
For vectors $v_1,\cdots,v_n\in\R^n$, the following identity holds
$$
||v_1+\cdots+v_n||^2=\displaystyle\sum_{i=1}^n||v_i||^2+2\displaystyle\sum_{1\le i<j\le n}v_i\cdot v_j.
$$
\proof
First consider the following identity from the definition of the norm for vectors $u,v\in\R^n$:
$$
||u+v||^2=(u+v)\cdot (u+v)=u\cdot u+u\cdot v+ v\cdot u + v\cdot v=||u||^2+||v||^2+2u\cdot v.
$$
Now if we let $u=v_1$ and $v= v_2+\cdots+v_n$ we get 
$$
||v_1+v_2+\cdots+v_n||^2=||v_1||^2+||v_2+\cdots +v_n||^2+2\sum_{i=2}^n v_1\cdot v_i.
$$
If we recurse on $||v_2+\cdots +v_n||^2$ using induction we get the desired result. 
\end{theorem}

\section{$N = n+1$}
Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$ in this case.

@@

\begin{figure}[!h]
    \centering
        \begin{tikzpicture}[yscale=-1, scale=0.75]
        % x-axis
        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
        % y-axis
        \draw [thick,->] (0, 4.5) -- (0, -4.5);
        % z-axis
        \draw [thick, ->] (3.0, -2.0) -- (-3.0, 2.0);
        % origin label
        \node at (-1, -0.3) {\text{$(0, 0, 0)$}};
        % x-axis label
        \node at (4.5, 0.5) {\text{$x$}};
        % y-axis label
        \node at (0, -5) {\text{$y$}};
         % z-axis label
        \node at (-3.4, 1.9) {\text{$z$}};
        \node at (3.5, -0.4) {\text{$(1, 0, 0)$}};
        \node at (1, -3.4) {\text{$(0, 1, 0)$}};
        \node at (-2.8, 1.1) {\text{$(0, 0, 1)$}};
        % circle
        \draw (3,0)[blue,fill=blue] circle (0.1cm);
        \draw (0, -3)[blue,fill=blue] circle (0.1cm);
        \draw (-1.8, 1.2)[blue,fill=blue] circle (0.1cm);

        \draw [thick,-,red,dashed] (3, 0) -- (0, -3);
        \draw [thick,-,red,dashed] (0, -3) -- (-1.8, 1.2);
        \draw [thick,-,red,dashed] (3, 0) -- (-1.8, 1.2);

        \draw [thick,-,blue] (3, 0) -- (0, 0);
        \draw [thick,-,blue] (0, -3) -- (0, 0);
        \draw [thick,-,blue] (-1.8, 1.2) -- (0, 0);
    \end{tikzpicture}
    \begin{tikzpicture}[yscale=-1, scale=0.75]
        % x-axis
        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
        % y-axis
        \draw [thick,->] (0, 4.5) -- (0, -4.5);
        % origin label
        \node at (-0.6, 0.3) {\text{$(0, 0)$}};
        % x-axis label
        \node at (4.5, 0.5) {\text{$x$}};
        % y-axis label
        \node at (0, -5) {\text{$y$}};
        % circle
        \draw (0,0) circle (3cm);
        \draw (3,0)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, 2.59)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);

@@

        \draw [thick,-,blue] (3, 0) -- (0, 0);
        \draw [thick,-,blue] (-1.5, 2.59) -- (0, 0);
        \draw [thick,-,blue] (-1.5, -2.59) -- (0, 0);
    \end{tikzpicture}

    \caption{A 2-dimensional simplex drawn in $\R^3$ on the left, and then centered at the origin in $\R^2$ on the right.}
\end{figure}

@@

We must first describe the vectors that make up the origin centered simplex. To center our simplex at the origin we take each vector in the original simplex, and subtract from it the centroid of the original simplex. The centroid of the original simplex takes the form $(\frac{1}{n},\frac{1}{n},\cdots,\frac{1}{N})$, so the points of the new simplex take the form
$$
\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
If we renormalize these vectors so that each vector has unit length, we get 
$$
\sqrt{\frac{N}{N-1}}\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
If we take the dot product of any two such vectors we get
$$
\frac{N}{N-1}\left(-2\cdot\frac{N-1}{N^2}+(N-2)\cdot\frac{1}{N^2}\right)=\frac{N}{N-1}\cdot-\frac{1}{N}=-\frac{1}{N-1}=-\frac{1}{n}.
$$
So an upper bound on the value of $\delta$ for all configurations of $n+1$ vectors in $n+1$ dimensional space is $\delta\le\frac{-1}{n}$.

@@

$$
||v_1+\cdots+v_{n+1}||^2=\displaystyle\sum_{i=1}^{n+1}||v_i||^2+2\displaystyle\sum_{1\le i<j\le n+1}v_i\cdot v_j.
$$
In our case we know three things. First, since the vectors are arranged in a simplex with centroid as the origin, $v_1+\cdots+v_{n+1}=0$. Second, $v_i\cdot v_j\le\delta$ for any $i<j$. Third, $||v_i||=1$. Therefore 
$$
0\le n+1+n(n+1)\delta.
$$
This implies that $\delta\ge -\frac{1}{n}$ as desired. Now we see that $\delta=-\frac{1}{n}$, so $|\delta|=\frac{1}{n}$, and therefore $\epsilon=\frac{1}{n^2}$.

\section{$N>n+1$}

We conjecture that in cases where $N>n+1$ we can never have an arrangement of the vectors where all values of $|v_i\cdot v_j|^2$ are the same. Proving this remains an object of future study.
